---
title: "OA paper"
author: "many"
output: github_document
---

*It's very much a work in progress; feel free to submit pull requests with updates and changes.*

```{r setup, echo=FALSE, warning=FALSE, message=FALSE, results='hide'}
library(knitr)
opts_knit$set(progress=TRUE, tidy=TRUE, echo=TRUE)
# devtools::install_github("cboettig/knitcitations@v1")
library(knitcitations); cleanbib()
cite_options(citation_format = "pandoc", check.entries=FALSE)

library(bibtex)
library(magrittr)
library(ggplot2)
library(dplyr)
library(broom)
library(reshape)
library(scales)
library(corrplot)
library(forcats)
library(jsonlite)
library(tidyr)
```

```{r}
2+2
```

```{r get_data, echo=FALSE, cache=TRUE, results='hide'}
# setwd("~/Dropbox/ti/oadoi-paper")

#### uncomment this on first run of the day
articles_raw <- read.csv("export_study_hybrid_20170504.csv")
articles_all = articles_raw

# modify the colors to handle hybrid
articles_all = articles_all %>%
  mutate(is_open_license=(grepl("cc", license) | oa_color=="gold"))

articles_all$oa = "closed"
articles_all$oa[articles_all$oa_color=="green"] = "green_only"
articles_all$oa[articles_all$oa_color=="gold"] = "gold_doaj"
articles_all$oa[articles_all$oa_color=="blue" & articles_all$is_open_license] = "gold_not_doaj"
articles_all$oa[articles_all$oa_color=="blue" & !articles_all$is_open_license] = "free"



# sort the factor for easier plotting
articles_all = mutate(articles_all, oa=fct_infreq(oa))

# how does it look
articles_all %>% select(oa_color, oa, is_open_license)
articles_all %>% count(oa)



# handy for color-coding later
oa_color_map = scale_fill_manual(values=c("#777777", "#2196F3", "#4CAF50", "#FFC107", "#FFEB3B"))


# articles_recent  %>% filter(journal=="Nature", oa_color=="blue") %>% select(oa, oa_color,evidence, license, year)


```





set up time frames
```{r} 


articles_timeseries = filter(articles_all, year >= 1990, year < 2016)  # just use last 50yrs
articles_recent = filter(articles_all, year >= 2009, year < 2016)

nrow(articles_all)
nrow(articles_timeseries)
nrow(articles_recent)
```

column names:
```{r}
colnames(articles_all)
```



```{r}

articles_all %>% select(evidence) %>% group_by(evidence) %>% tally()
```







overall oa picture
```{r}
ggplot(articles_all, aes(x="", fill=oa)) + geom_bar() + oa_color_map
```



growth in literature over time with gold/green showing. 
but this does not have percent
```{r}
ggplot(articles_timeseries, aes(x=year, fill=oa)) + geom_bar(width=1) + oa_color_map
```

position=fill makes it show percent. but ugly
we could chop it at 1990
```{r}
ggplot(articles_timeseries, aes(x=year, fill=oa)) + geom_bar(width=1, position="fill") + oa_color_map
```

another way to show percent
from http://stackoverflow.com/questions/24576515/relative-frequencies-proportions-with-dplyr
```{r}
oa_freq_by_year = articles_timeseries %>% count(year, oa) %>%  # count articles in each year and color
  mutate(perc = n / sum(n)) %>%  # add col for percent *within that year* cos we are still in Group Mode
  ungroup()  # get out of that Group Mode.

oa_freq_by_year %>% ggplot(aes(x=year, y=perc, fill=oa)) + geom_area() + oa_color_map
```



# publishers
# ========================

# sort the publisher factor by frequency...helps in plotting.
```{r}

articles_recent$publisher = fct_infreq(articles_recent$publisher)

publishers = articles_recent %>% 
  count(publisher, oa) %>%
  arrange(desc(n)) %>%
  ungroup()

# the top 25 publishers publish 80% of articles.
# top 100 publshers publish 87% of articles.
sum(publishers$n[0:100]) /sum(publishers$n)
```

```{r}
publishers %>% slice(1:25) %>% ggplot(aes(x=publisher, y=n)) + geom_bar(stat="identity") + coord_flip()
```

```{r}
publishers_oa = articles_recent %>% 
  count(publisher, oa) %>%
  ungroup()


publishers_oa %>% slice(1:80) %>% ggplot(aes(x=publisher, y=n, fill=oa)) + 
  geom_bar(stat="identity") + 
  coord_flip() +
  oa_color_map
```

same thing but by % oa
```{r}
publishers_oa %>% slice(1:80) %>% ggplot(aes(x=publisher, y=n, fill=oa)) + 
  geom_bar(stat="identity", position="fill") + 
  coord_flip() +
  oa_color_map

```





```{r}

articles_all$journal = fct_infreq(articles_all$journal)

journals = articles_all %>% 
  count(journal, oa) %>%
  arrange(desc(n)) %>%
  ungroup()

sum(journals$n[0:100]) /sum(journals$n)

journals_oa = articles_all %>% 
  count(journal, oa) %>%
  ungroup()


journals_oa %>% slice(1:275) %>% ggplot(aes(x=journal, y=n, fill=oa)) + 
  geom_bar(stat="identity") + 
  coord_flip() +
  oa_color_map
```

same thing but by % oa
```{r}
journals_oa %>% slice(0:76) %>% ggplot(aes(x=journal, y=n, fill=oa)) + 
  geom_bar(stat="identity", position="fill") + 
  coord_flip() +
  oa_color_map

```













_subjects_

unpack subjects into a list
```{r}
# articles %>% 
# select(subject, oa) %>%
# filter(as.character(subject) != "") %>%
# rowwise()  %>%
# mutate( subj_list = list( fromJSON( as.character( subject ) ) ) )


# articles$subject[0:10]
```



